\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap

### 2017-02-20 TeX/paper.tex
Update paper.tex

\begin{abstract}
With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, as the size of the network increases, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 10,000. When the network is large, one option for estimation is maximum pseudolikelihood (MPLE). MPLE is considerably less accurate than MLE in small networks, but exhibits comparable mean squared error in large networks. The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. Furthermore, bootstrapped MPLE has the advantage of being embarrassingly parallel. We compare the two different approaches by applying them on U.S. Supreme Court Citation Network Data.
\end{abstract}

@@

\section*{Efficiency of MPLE/MCMCMLE}
Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression analysis and does not exert elaborate MCMC methods. Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}). This means that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE should be considered as an option to computationally intensive MCMC methods.\\
On the other hand a major reason for MCMCMLE methods is, as shown by van Duijn et al. \cite{vanDuijn.2009}, that the MCMCMLE is more efficient than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE. However, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. Furthermore, the number of simulated networks needed in order for the MCMCMLE to surpass the MPLE increases as the size of the network increases, making it difficult to evaluate whether the chosen number of simulated networks is large enough. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the MCMCMLE approach.

\textcolor{blue}{We need at least two different networks of different size in order to show the points above.}\\
In order to proof these disadvantages of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students. The original data comes from Resnick et al \cite{Resnick.1997}.\\[0.3cm]
\begin{itemize}
\item explain simulation study
\end{itemize}
Based on these results we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 
$$RMSE = \sqrt{\sum_{i=1}^{n}(\theta - \hat{\theta})^2}$$
meaning that the smaller the RMSE, the more accurate the estimator. Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE, we take the log of the ratio of the MCMCMLE and the MPLE. As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance.

\begin{itemize}
\item RMSE plot + explanation + discussion
\end{itemize}

>>>>

\title{\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap}
\maketitle

\begin{abstract}
\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, as the size of the network increases, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 10,000. When the network is large, one option for estimation is maximum pseudolikelihood (MPLE). MPLE is considerably less accurate than MLE in small networks, but exhibits comparable mean squared error in large networks. The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. Furthermore, bootstrapped MPLE has the advantage of being embarrassingly parallel. We compare the two different approaches by applying them on U.S. Supreme Court Citation Network Data.
\end{abstract}

@@

\section*{Efficiency of MPLE/MCMCMLE}
Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods. Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
However, another major advantage of the MCMCMLE method is, as shown by van Duijn et al. \cite{vanDuijn.2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE. Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. It is even more remarkable that the number of simulated networks needed for the MCMCMLE, in order to surpass the MPLE increases as the size of the network increases. This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMCMLE to outperform the MPLE. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\

\noindent In order to demonstrate this disadvantage of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al \cite{Resnick.1997}.\\[0.3cm]
For both networks we first calculate the MCMCMLE and treat the estimated coefficients as the network's true values $\theta$. For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution. We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMCMLE of these simulated networks. For every single simulated network the MCMCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 
$$RMSE = \sqrt{\sum_{i=1}^{500}(\theta - \hat{\theta}_i)^2}$$
implying that the smaller the RMSE, the more accurate is the estimator. Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMCMLE to the MPLE. As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance. Figure \ref{rmse} visualizes the results of the simulation study. The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.5]{RMSE_paper}
\caption{The log of the ratio of the RMSE for the MCMCMLE to the MPLE for different samplesizes and two different networks, Faux Mesa High and Faux Magnolia High}
\label{rmse}
\end{center}
\end{figure}

\noindent The plots support the fact that larger networks require a larger samplesize of simulated networks in order for the MCMCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a samplesize of about 50 networks the larger Faux Magnolia High network already requires a samplesize of at least 1500 networks for the MCMCMLE to surpass the MPLE. These results propose especially for very large networks (e.g. social media data) the question, of how large the samplesize has to be set in order to justify the approximately exact, but comutationally expensive MCMCMLE method. The identification of the required samplesize is not only becoming more difficult as the size of the network increases, but an increase of the samplesize will also extends the mere computation time.
